---
title: "nomogrammer: Fagan's nomogram using ggplot2"
author: "Adam M. Chekroud"
date: "12/31/2016"
output:
    html_document:
        theme: flatly
        toc: true
        toc_depth: 2

---

Tweet me @[itschekkers](https://twitter.com/itschekkers)!

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev="svg", fig.width=4, fig.height=4, fig.align="center")
```

<!--html_preserve-->
<meta name="twitter:card" content="summary_large_image">
<meta name="twitter:site" content="@itschekkers">
<meta name="twitter:creator" content="@itschekkers">
<meta name="twitter:title" content="nomogrammer: simple nomogram plots in R using ggplot2">
<meta name="twitter:description" content="nomograms are an easy way of calculating post-test probabilities for a diagnostic test. Now you can create them in two lines of R code!">
<!--/html_preserve-->

```{r, echo=FALSE}
source("https://raw.githubusercontent.com/achekroud/nomogrammer/master/nomogrammer.r")
```

# Introduction

The `nomogrammer` function is a simple visualization function to plot Fagan's nomograms as [`ggplot2`][ggplot2] objects. 

They are used to illustrate the change from pre-test to post-test probabilities that occur when a predictive test is used.

The underlying calculation is as follows:

- $odds = p / (1-p)$ 
- By Bayes' rule :    $posttest (posterior) odds = likelihood ratio * pretest (prior) odds$ 
- Taking logs :    $log(posterior) = log(prior) + log(LR)$


## Rationale

The [Fagan nomogram](http://www.pmean.com/definitions/fagan.htm) is a graphical tool for estimating how much the result on a diagnostic test changes the probability that a patient has a disease (NEJM 1975; 293: 257). Although there are R packages for creating more complicated nomograms (e.g. the `rms` [package](https://cran.r-project.org/web/packages/rms/rms.pdf), and web-implementations for creating simple Fagan nomograms, there is no simple nomogram plotting function in R. 

The `nomogrammer` function offers such a plotting method, using the "grammar of graphics" implemented in the [`ggplot2`][ggplot2] package to render the plot. Its output is designed to be close to that which is normally obtained through [online nomogram builders](https://araw.mede.uic.edu/cgi-bin/testcalc.pl). 

## Installation

`nomogrammer` is a standalone function:

```{r, eval=FALSE}
source("https://raw.githubusercontent.com/achekroud/nomogrammer/master/nomogrammer.r")
```

## Dependencies

There are two main package dependencies for `nomogrammer`:  [`ggplot2`][ggplot2] and [`scales`][scales]

```{r, message=FALSE, eval = F}
library(ggplot2)
library(scales)
```

The [`ggplot2`][ggplot2] package can be installed from [CRAN](https://cran.r-project.org/) through [`install.packages`](http://www.rdocumentation.org/packages/utils/functions/install.packages). Do the same to install the scales package, i.e. `install.packages("scales")`

# Basic Example

The function always requires the prevalence (i.e. prior probability), expressed as a probability (not a percentage).

You then either give it the positive and negative likelihood ratios:

```{r, eval=FALSE}
nomogrammer(Prevalence = .60,
            Plr = 12,
            Nlr = 0.6)
```

Or the sensitivity and specificity of the model.

```{r, eval=FALSE}
nomogrammer(Prevalence = .60,
            Sens = 0.421,
            Spec = 0.965)
```

And it will plot the basic nomogram

```{r, echo = FALSE, eval=TRUE, message = FALSE, warning=FALSE}
nomogrammer(Prevalence = .60,
            Sens = 0.421,
            Spec = 0.965)
```

The nomogram is returned as a ggplot object, which you could save and plot as usual

```{r, eval=F}
p <- nomogrammer(Prevalence = .60, Sens = 0.421, Spec = 0.965)
plot(p)
```

# Options

There are a couple of optional extras with this function.

You can use the `NullLine` argument to overlay a line that goes from your prior probability through LR = 1, illustrating the posterior probability if it remained unchanged.

```{r, echo = FALSE, eval=TRUE}
nomogrammer(Prevalence = .60,
            Sens = 0.421,
            Spec = 0.965,
            NullLine = TRUE)
```

You can use the `Detail` argument to overlay some helpful metrics in the top right corner.

```{r, echo = FALSE, eval=TRUE, fig.width=6, fig.height=4}
nomogrammer(Prevalence = .60,
            Sens = 0.421,
            Spec = 0.965,
            Detail = TRUE)
```


# Visual tweaks using `ggplot` methods

Since `nomogrammer` returns a `ggplot` object, it is easy to edit with usual methods

You can add a title, for example

```{r, eval=T}
p <- nomogrammer(Prevalence = .60, Sens = 0.421, Spec = 0.965)
p + ggtitle("put your title here", subtitle = "and your subtitle here")
```

You can change the colours of the lines, either the lazy way (not run)

```{r, eval=FALSE}
## Not run
nomogrammer(Prevalence = .60, Plr = 120, Nlr = 0.4) +
    scale_color_manual(values=c("pink", "black"))
```

Or the precise way (hex codes):

```{r, eval=T}
nomogrammer(Prevalence = .60, Plr = 120, Nlr = 0.4) +
    scale_color_manual(values=c("#ff7f00", "#b2df8a"))
```

In general, further control over plot is possible through any  [`theme`](http://docs.ggplot2.org/current/theme.html) and [`guides`](http://docs.ggplot2.org/current/guides.html) methods that also apply:

# Saving

Saving the plot is also easy using the `ggsave` function (here I picked some size settings that looked reasonable):

```{r, echo = TRUE, eval=F}
p <- nomogrammer(Prevalence = .60, Sens = 0.421, Spec = 0.965)

ggsave(p, file = "~/nomogrammer/plot.pdf",
       width = 5.99,
       height = 3.99, units = "in")
```

And, of course, you can save the plot however you like. Here it will save as a png instead:

```{r, echo = TRUE, eval=F}
demo <- nomogrammer(Prevalence = .60, Sens = 0.66, Spec=0.77) +
            ggtitle("Fagan's nomogram",
                    subtitle = "Example nomogram made with this function")

ggsave(demo, file = "~/nomogrammer/demo.png",
       width = 5.99,
       height = 3.99, units = "in")
```



PS You can also use the `Verbose` argument if, for whatever reason, you want the function to return the main statistics to the console.

__Note:__ this doesn't work well in Rmd, so isn't run here.

```{r, echo = T, eval=F}
# Not run:
# > nomogrammer(Prevalence = .60, Sens = 0.66, Spec=0.77, Verbose = TRUE)

# Returns:
# prevalence = 60%
# PLR = 2.87
# NLR = 0.442
# posterior probability (positive) = 81.1%
# posterior probability (negative) = 39.8%
# sensitivity = 66%
# specificity = 77%
```




# Known limitations

`nomogrammer` is strictly limited to plotting Fagan's nomograms quickly: it cannot plot complex nomograms that visualize multiple covariates, or non-linearities. Other packages exist for this (e.g. pynomo, rms, or hdnom)

If you find other limitations or would like to contribute improvements or corrections to `nomogrammer`, please feel free to submit a pull request or [submit an issue](https://github.com/achekroud/nomogrammer/issues), thanks!

Tweet me @[itschekkers](https://twitter.com/itschekkers)!


---

[colorbrewer]: http://colorbrewer2.org/
[ggplot2]: http://ggplot2.org/
[scales]: http://www.rdocumentation.org/packages/scales

